%%-*-latex-*-

\section{Naive approach}
\label{naive}
 
\noindent Given a query \(\query{\kappa_1}{\kappa_2, \dots,
  \kappa_{{\cal A}(q)}}\), a naive approach to solve our problem
consists in storing the elements read so far in lists according to
their kind (as in the multiple streams framework) and, when a new
element is read, we check for all the combinations of elements that
make up a match. This procedure is inefficient because it reconsiders
previous invalid combinations after an element is read. We can easily
improve it by storing all the partial matches found so far and then
trying to complete \emph{only them} with one more element each time,
until a complete match is found. For example, let us consider the
stream [\(d_1\), \(a_1\), \(c_1\), \(b_1\), \(c_2\), \(a_2\), \(b_2\)]
and let us seek both ordered and unordered matches. In
figure~\ref{input_lists}, the first line corresponds to the
\begin{wrapfigure}[8]{r}{0pt}
\centering
\fbox{\(
\begin{array}{c|c|c|c|c|c|c}
\varnothing & a_1 & a_1 & a_1 & a_1 & a_1,a_2 & a_1, a_2\\
\varnothing & \varnothing & \varnothing & b_1 & b_1 & b_1 & b_1,b_2\\
\varnothing & \varnothing & c_1 & c_1 & c_1,c_2 & c_1,c_2 & c_1,c_2\\
d_1 & d_1 & d_1 & d_1 & d_1 & d_1 & d_1
\end{array}
\)
}
\caption{Input lists.\label{input_lists}}
\end{wrapfigure}
sorted list of \(a\)-elements read so far, the second to the
\(b\)-elements etc. Each column shows the state of the lists after
each element is read from the input stream.  At each step, i.e., for
each column, the new valid combinations are respectively
\(\varnothing\), \{\(a_1\)\}, \{\(a_1\), \(a_1c_1\)\}, \{\(a_1\),
\(a_1c_1\), \(a_1b_1\)\}, \{\(a_1\), \(a_1c_1\), \(a_1b_1\),
\(\mathbf{a_1b_1c_2}\), \(a_1c_2\)\}, \{\(a_1\), \(a_1c_1\),
\(a_1b_1\), \(\mathbf{a_1b_1c_2}\), \(a_1c_2\), \(a_2\)\}, \{\(a_1\),
\(a_1c_1\), \(a_1b_1\), \(\mathbf{a_1b_1c_2}\), \(a_2\), \(a_1c_2\),
\(\mathbf{a_1b_2c_1}\), \(a_1b_2\)\} ---~solutions in bold. Note that,
since the kinds in this query are pairwise distinct, we can simplify
the notation for complete and partial matches by just enumerating
elements, like \(a_1c_1\) and \(a_1b_2\) etc. The drawbacks of this
procedure are that it uses too much memory and it is too slow. Indeed,
it needs to keep in memory the complete lists of elements and it tries
to form new combinations with elements which are not disjoint, like
\(a_1b_1c_1\) in our example (\(b_1 \sqsubset c_1\)).
